---
title: "101220_Rubangisa"
author: "Wayne Rubangisa"
date: "2025-05-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# LIFE EXPECTANCY ANALYSIS

## Download and load data to R Studio

### Country Population

```{r}
country_population <- read.csv("country_population.csv")
head(country_population)
```
### Fertility Rate

```{r}
fertility_rate <- read.csv("fertility_rate.csv")
head(fertility_rate)
```

### Life Expectancy

```{r}
life_expectancy <- read.csv("life_expectancy.csv")
head(life_expectancy)
```

### Life Expectancy Data

```{r}
life_expectancy_data <- read.csv("Life Expectancy Data.csv")
head(life_expectancy_data)
```

## Exploratory Data Analysis

### Display Variable Names

See the names of all variables in the data set `Life Expectancy Data`

```{r}
names(life_expectancy_data)
```


```{r}
cat(names(life_expectancy_data), sep = "\n")
```

### Display the first 5 rows

Preview the first 5 observations in the data set. 

```{r}
head(life_expectancy_data, 5)
```

### Display the bottom 10 rows

View the last 10 observations of the data set.

```{r}
tail(life_expectancy_data, 10)
```
### Display the data types

Shows what type of data each column holds.

```{r}
sapply(life_expectancy_data, class)
```

We need a cleaner way to present this though. 

```{r}
result <- sapply(life_expectancy_data, class)
result
```

```{r}
result <- sapply(life_expectancy_data, class)
data.frame(Column = names(result), Class = result)
```
Let's see if we can remove the extra (the first column). 

```{r}
result <- cat(sapply(life_expectancy_data, class), sep = "\n")
result
```


```{r}
names(result)
```
We still need the names though. So we need to find a way to get the names and values. 

```{r}
result
```

```{r}
result[]
```
```{r}
class(result)
```
```{r}
unname(result)
```
```{r}
data.frame(Column = names(result), Class = unname(result))

```

```{r}
sapply(life_expectancy_data, class)
```
It checks out. We can move to the next tasks. 

### Shape of the dataset.

```{r}
dim(life_expectancy_data)
```

```{r}
cat("Rows:", dim(life_expectancy_data)[1], "\n")
cat("Columns:", dim(life_expectancy_data)[2])
```
Another way: 

```{r}
rows <- nrow(life_expectancy_data)
cols <- ncol(life_expectancy_data)
cat("Rows:", rows, "\n")
cat("Columns:", cols)
```
### Drop Duplicate Rows

First of all, let's see how many duplicated observations we have in our dataset. 

`duplicated(life_expectancy_data)` returns a logical vector. Instead of printing the vector, we want to know how many trues we have in said vector. 

```{r}
sum(duplicated(life_expectancy_data))
```
Well, it looks like we don't need to drop duplicates, since there aren't any. In which case, we move on to the next task. 

### Number of missing values in each column

```{r}
colSums(is.na(life_expectancy_data))
```

Now, let's make our output presentable. 

```{r}
data.frame(Columns = names(colSums(is.na(life_expectancy_data))), class = unname(colSums(is.na(life_expectancy_data)))) 
```

```{r}
library(dplyr)
data.frame(Columns = names(colSums(is.na(life_expectancy_data))), class = unname(colSums(is.na(life_expectancy_data)))) %>%
  arrange(desc(class))
```

### Use boxplots to check for outliers in quantitative variables. 

```{r}
numeric_vars <- sapply(life_expectancy_data, is.numeric)
boxplot(life_expectancy_data[, numeric_vars], main = "Boxplots for Numeric Variables", las = 2)
```

The boxplots are a bit small though. Let's try to make them bigger. 

```{r}
boxplot(life_expectancy_data[, numeric_vars], main = "Boxplots for Numeric Variables", las = 2, boxwex = 2.5)
```

`boxwex` didn't work. Let's try another one. 

```{r}
library(reshape2)
long_data <- melt(life_expectancy_data[, numeric_vars])

library(ggplot2)
ggplot(long_data, aes(x = variable, y = value)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  ggtitle("Boxplots for Numeric Variables")

```



```{r}
# Create a ggplot boxplot with more control over size
ggplot(long_data, aes(x = variable, y = value)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),  # Rotate labels
        plot.margin = margin(10, 10, 10, 10),  # Adjust plot margins
        axis.text = element_text(size = 12),  # Change axis text size
        axis.title = element_text(size = 14)) +  # Change title size
  ggtitle("Boxplots for Numeric Variables")
```

### Handle missing values and outliers. 

Which columns are continuous so that we can fill na with the media (since we have established that we have outliers).

```{r}
continuous_cols <- colnames(life_expectancy_data)[sapply(life_expectancy_data, is.numeric)]
continuous_cols
```

Now, we replace the NA values in each of the these columns by the median. Now, let's handle some missing values. 

```{r}
for (col in continuous_cols) {
  life_expectancy_data[[col]] <- ifelse(
    is.na(life_expectancy_data[[col]]), 
    median(life_expectancy_data[[col]], na.rm = TRUE), life_expectancy_data[[col]])
}
```

Now, let's see if we have any missing values. 

```{r}
library(dplyr)
data.frame(Columns = names(colSums(is.na(life_expectancy_data))), class = unname(colSums(is.na(life_expectancy_data)))) %>%
  arrange(desc(class))
```
Wonderful. Now, let's handle some outliers. This is fun. 🤩

We are going to use the Interquartile Range (IQR) method to identify and cap or remove outliers. 

```{r}
cap_outliers <- function(x) {
  q1 <- quantile(x, 0.25)
  q3 <- quantile(x, 0.75)
  iqr <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr
  x[x < lower] <- lower
  x[x > upper] <- upper
  return(x)
}

life_expectancy_data[numeric_vars] <- lapply(life_expectancy_data[numeric_vars], cap_outliers)

```


Let's see if we still have outliers. 

```{r}
boxplot(life_expectancy_data[, numeric_vars], main = "Boxplots for Numeric Variables", las = 2)
```
Excellent. 

## Value Extraction and Plot

### Filter for 2015 and get top 10 countries by population
We’ll find which 10 countries had the highest population in 2015.

```{r}
head(country_population)
```

Let’s get the top 10 countries by 2015 population:

```{r}
head(population_2016 <- country_population %>% select(`Country.Name`, `X2016`))
```
Let's see if we have any missing values. 
```{r}
colSums(is.na(population_2016))
```
Let's remove them. 

```{r}
population_2016 <- na.omit(population_2016)
colSums(is.na(population_2016))
```
Get top 10 countries by 2016 population

```{r}
top10_2016 <- population_2016 %>%
  arrange(desc(`X2016`)) %>%
  slice(1:10)
top10_2016
```
Wait, we getting entries that are not countries. Let's fix that. 

```{r}
# Define a vector of known non-country entries
non_country_entries <- c("World", "IDA & IBRD total", "Low & middle income", "Middle income",
                         "IBRD only", "Early-demographic dividend", "Lower middle income",
                         "Upper middle income", "East Asia & Pacific", "Late-demographic dividend")

# Filter out non-country entries
countries_only <- population_2016 %>%
  filter(!`Country.Name` %in% non_country_entries)

# Select top 10 countries by population in 2016
top10_2016 <- countries_only %>%
  arrange(desc(`X2016`)) %>%
  slice(1:10)

top10_2016
```

Still needs more fixing. We'll use the country code and only select the observations with valid country codes. 

```{r}
population_2016 <- country_population %>% select(`Country.Name`, `Country.Code`, `X2016`)
head(population_2016)
```

```{r}
valid_iso_codes <- c(
  "AFG", "ALB", "DZA", "AND", "AGO", "ARG", "ARM", "AUS", "AUT", "AZE",
  "BDI", "BEL", "BEN", "BFA", "BGD", "BGR", "BHR", "BIH", "BLR", "BLZ", 
  "BOL", "BRA", "BRN", "BTN", "BWA", "CAF", "CAN", "CHE", "CHL", "CHN",
  "CIV", "CMR", "COD", "COG", "COL", "COM", "CPV", "CRI", "CUB", "CYP",
  "CZE", "DEU", "DJI", "DNK", "DOM", "DZA", "ECU", "EGY", "ERI", "ESP",
  "EST", "ETH", "FIN", "FJI", "FRA", "GAB", "GBR", "GEO", "GHA", "GIN",
  "GMB", "GNB", "GNQ", "GRC", "GTM", "GUY", "HND", "HRV", "HTI", "HUN",
  "IDN", "IND", "IRL", "IRN", "IRQ", "ISL", "ISR", "ITA", "JAM", "JOR",
  "JPN", "KAZ", "KEN", "KGZ", "KHM", "KIR", "KOR", "KWT", "LAO", "LBN",
  "LBR", "LBY", "LKA", "LSO", "LTU", "LUX", "LVA", "MAR", "MDA", "MDG",
  "MDV", "MEX", "MKD", "MLI", "MLT", "MMR", "MNE", "MNG", "MOZ", "MRT",
  "MUS", "MWI", "MYS", "NAM", "NER", "NGA", "NIC", "NLD", "NOR", "NPL",
  "NZL", "OMN", "PAK", "PAN", "PER", "PHL", "PNG", "POL", "PRT", "PRY",
  "QAT", "ROU", "RUS", "RWA", "SAU", "SDN", "SEN", "SGP", "SLB", "SLE",
  "SLV", "SOM", "SRB", "SSD", "SUR", "SVK", "SVN", "SWE", "SWZ", "SYR",
  "TCD", "TGO", "THA", "TJK", "TKM", "TLS", "TTO", "TUN", "TUR", "TZA",
  "UGA", "UKR", "URY", "USA", "UZB", "VEN", "VNM", "VUT", "YEM", "ZAF",
  "ZMB", "ZWE"
)

# Filter only valid countries using Country.Code
countries_only <- population_2016 %>%
  filter(Country.Code %in% valid_iso_codes)

# Get top 10 by population in 2016
top10_2016 <- countries_only %>%
  arrange(desc(`X2016`)) %>%
  slice(1:10) %>% select(Country.Name, `X2016`)

top10_2016 

```
Beautiful, isn't it. Let's move on. 

### Visualize with ggplot2
We'll use a bar chart, which is appropriate for comparing numerical values across categories (countries, in our case).

Plot 2016 population of those top 10 countries

```{r}

ggplot(top10_2016, aes(x = reorder(`Country.Name`, `X2016`), y = `X2016`)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  # Flips axes for better readability
  labs(title = "Top 10 Most Populous Countries in 2015 and Their 2016 Population",
       x = "Country",
       y = "Population (2016)") +
  theme_minimal()
```

```{r}

library(tidyr)
library(readr)
library(stringr)
top_10_countries <- top10_2016$Country.Name

fertility_top <- fertility_rate %>%
  filter(`Country.Name` %in% top_10_countries)

# STEP 3: Reshape to long format
fertility_long <- fertility_top %>%
  pivot_longer(
    cols = `X1960`:`X2016`,
    names_to = "Year",
    values_to = "Fertility_Rate"
  ) %>%
  mutate(
    Year = as.numeric(str_remove(Year, "^X"))
  )

# STEP 4: Plot the trend using ggplot2
ggplot(fertility_long, aes(x = Year, y = Fertility_Rate, color = `Country.Name`)) +
  geom_line(size = 1.2) +
  labs(
    title = "Fertility Rate Trends (1960–2016)",
    x = "Year",
    y = "Fertility Rate (births per woman)",
    color = "Country"
  ) +
  theme_minimal()

```

```{r}
life_expectancy_top <- life_expectancy %>%
  filter(`Country.Name` %in% top_10_countries)

# STEP 3: Reshape to long format (wide ➝ long)
life_expectancy_long <- life_expectancy_top %>%
  pivot_longer(
    cols = `X1960`:`X2016`,
    names_to = "Year",
    values_to = "Life_Expectancy"
  ) %>%
  mutate(
    Year = as.numeric(str_remove(Year, "^X")),
    Life_Expectancy = as.numeric(Life_Expectancy)
  )

# STEP 4: Plot the life expectancy trend
ggplot(life_expectancy_long, aes(x = Year, y = Life_Expectancy, color = `Country.Name`)) +
  geom_line(size = 1.2) +
  labs(
    title = "Life Expectancy Trends (1960–2016)",
    x = "Year",
    y = "Life Expectancy (Years)",
    color = "Country"
  ) +
  theme_minimal()
```

```{r}

# Selecting only the columns needed for correlation
cols_to_use <- c("Life.expectancy", "Adult.Mortality", "infant.deaths", "Alcohol",
                 "percentage.expenditure", "Hepatitis.B", "Measles", "BMI",
                 "under.five.deaths", "Polio", "Total.expenditure", "Diphtheria",
                 "HIV.AIDS", "GDP", "Population", "thinness..1.19.years",
                 "thinness.5.9.years", "Income.composition.of.resources", "Schooling")

# Subset and remove rows with missing values
data_numeric <- life_expectancy_data[, cols_to_use]
data_numeric <- na.omit(data_numeric)  # Removes rows with any NA values
```

```{r}
# Calculate correlation matrix using Pearson method (default)
cor_matrix <- cor(data_numeric)

# View the first few rows of correlation matrix
head(cor_matrix)
```

```{r}
library(ggcorrplot)

ggcorrplot(cor_matrix, 
           lab = TRUE, 
           lab_size = 2.5, 
           method = "square", 
           colors = c("red", "white", "blue"), 
           title = "Correlation Matrix", 
           hc.order = TRUE, 
           type = "lower")
```
```{r}

```

```{r}
cor(data_numeric$Life.expectancy, data_numeric$Schooling)
```
This indicates a strong positive correlation between life expectancy and schooling.

Interpretation:
A value of 0.73 means that as schooling increases, life expectancy tends to increase as well.

Educated individuals are likely more aware of health risks, make better lifestyle choices, and have access to better jobs and healthcare.

This relationship makes logical sense and suggests that improving education might indirectly improve public health and longevity.



```{r}
# Check column names of both datasets
names(life_expectancy_data)            # Life expectancy dataset
names(fertility_rate)  # Fertility dataset

# Rename columns if needed for consistency (assume they are "Country", "Year", "FertilityRate")
colnames(fertility_rate)[colnames(fertility_rate) == "FertilityRate"] <- "fertility_rate"
```


```{r}
fertility_long <- fertility_rate %>%
  pivot_longer(
    cols = starts_with("X"),  # All year columns like X1960, X1961...
    names_to = "Year",
    values_to = "fertility_rate"
  )


fertility_long

# Check structure
```

```{r}
# Load libraries
library(tidyverse)
library(caret)      # For RMSE
library(Metrics)    # For RMSE (optional)
library(psych)      # For describe()
library(ggplot2)

# Fit the linear regression model
model <- lm(Life.expectancy ~ ., data = life_expectancy_data)

# View model summary
summary(model)

```

```{r}
# Install if not installed already
library(Metrics)
# Predict life expectancy using the model
predictions <- predict(model, newdata = life_expectancy_data)

# Actual values
actual <- life_expectancy_data$Life.expectancy

# Calculate RMSE
rmse_value <- rmse(predictions, actual)

# Print RMSE and R-squared
cat("R-squared:", summary(model)$r.squared, "\n")
cat("RMSE:", rmse_value, "\n")
```
```{r}
# Define EAC countries
eac_countries <- c("Burundi", "Kenya", "Rwanda", "South Sudan", "Tanzania", "Uganda")

# Define SADC countries
sadc_countries <- c("Angola", "Botswana", "Comoros", "Congo, Dem. Rep.",
                    "Eswatini", "Lesotho", "Madagascar", "Malawi", "Mauritius",
                    "Mozambique", "Namibia", "Seychelles", "South Africa",
                    "Tanzania", "Zambia", "Zimbabwe")
```

```{r}
# Extract EAC and SADC subsets
eac_data <- life_expectancy_data %>%
  filter(Country %in% eac_countries & !is.na(Life.expectancy)) %>%
  mutate(Region = "EAC")

sadc_data <- life_expectancy_data %>%
  filter(Country %in% sadc_countries & !is.na(Life.expectancy)) %>%
  mutate(Region = "SADC")
```

```{r}
# Combine the two datasets
eac_sadc_data <- bind_rows(eac_data, sadc_data)

# Group by region and calculate average life expectancy
region_life_exp <- eac_sadc_data %>%
  group_by(Region) %>%
  summarise(Average_Life_Expectancy = mean(Life.expectancy))

# View results
print(region_life_exp)
```

```{r}
# Bar plot for comparison
ggplot(region_life_exp, aes(x = Region, y = Average_Life_Expectancy, fill = Region)) +
  geom_col() +
  labs(title = "Average Life Expectancy in EAC vs SADC (2013)",
       x = "Region",
       y = "Average Life Expectancy (Years)") +
  theme_minimal() +
  theme(legend.position = "none")
```

